---
title: "Add project title"
author: "Max Niu"
date: "Update date"
format: pdf
execute: 
  warning: false
  message: false
  echo: false
editor: visual
---

```{r}
#| label: load-pkg-data
library(tidyverse)
library(tidymodels)
library(readxl)
library(janitor)
library(lubridate)
library(knitr)

raw_events <- read_excel("data/emdat_disasters_2000_2025.xlsx") 

```

## Data processing

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
events <- raw_events |>
  clean_names()

events <- events |>
  select(disaster_type, country, start_year, start_month, start_day, end_year, end_month, 
         end_day, total_deaths, no_injured, no_affected, total_affected, total_damage_000_us)


events <- events |>
  drop_na() #selecting was done before because otherwise too many rows would be dropped

events <- events |>
  mutate(
    start_date = make_date(start_year, start_month, start_day),
    end_date = make_date(end_year, end_month, end_day),
    duration_days = as.numeric(end_date - start_date)
  )

```

##  Introduction

#### Background:

As global temperatures rise, climate-related disasters have increased in both frequency and severity over the past several decades ([World Wild Life](https://www.worldwildlife.org/resources/explainers/is-climate-change-increasing-the-risk-of-disasters/)). Extreme weather events such as floods, droughts, hurricanes, etc. impose a huge human and real economic costs across the world. It is imperative that we understand the determinants of disaster severity (in terms of economic damages) for better policy planning and mitigative strategies.

The Emergency Events Database (EM-DAT), maintained by the Centre for Research on the Epidemiology of Disasters (CRED), keeps an internationally standardized data on natural diasters worldwide. The information includes disaster type, location, affected population, deaths, injuries, and economic damages.

#### Research Question:

How do different disaster characteristics (such as geographic location, time of year, type, and total casualties incurred) predict the total economic damage caused by natural disasters in USD?

-   Specifically, we examine whether disaster type, location, timing, duration, and measures of human impact (deaths, injuries, and affected population) are associated with higher reported economic losses.

#### Motivation:

We were first drawn to this research question because were intrigued by this data; we were amazed by the granularity and the surprising breadth of information in this data. However, we realized that doing research on this matter could have real, practial uses.

First, identifying predictors of large economic damage is critical for both public and private institutions. Governments and international organizations allocate substantial resources towards mitigation and recovery efforts, yet the determinants of high-cost disasters vary across event types and regions. Same applies for insurance companies and hedge fund traders that try modeling key determinants relevant to a disaster for their comparative advavantage against peers in the market.

By analyzing global disaster data, this project aims to quantify which observable characteristics are most strongly associated with economic loss. We hope that these results may provide insight into which types of disasters and impact measures are most economically destructive.

#### Hypothesis Reasoning:

Based on economic reasoning and relevant research, we expect that...

1.  Disasters with higher number of affected individuals (absolute number, number of deaths and injuries) will have a positive relationship with economic damage.
2.  Certain disaster types will incur systematically higher economic damage than that of others.
3.  The level of damage dollars will defer based on countries because of their respective infrastructure that exists to mitigate damage.

#### Data Description:

This data was obtained from the Emergency Events Database, otherwise known as [EM-DAT](https://doc.emdat.be/docs/data-structure-and-content/emdat-public-table/#column-description), which is maintained by the Centre for Research on the Epidemiology of Disasters (CRED). The goal of the database is to compile standardized information on natural disasters that occur around the world. According to the information on metadata, an event is included in EM-DAT if one or more of the following criteria is met:

-   10+ people killed
-   100+ people affected
-   Declaration of a state of emergency
-   Solicitation of international assistance

The observations describe different disasters that occurred around the world. Of each disaster, data on the number of people injured, affected, homeless, or killed, the total economic damages, type of disaster, exact location, start date and end date of the disaster were collected.

The data dictionary can be found [here](https://github.com/sta210-sp26/project-MaxTGH-hyeonsuan06/blob/main/data/README.md).

## Exploratory Data Analysis

```{r}
events |> #with outliers
  ggplot(aes(x = total_damage_000_us)) +
  geom_histogram() + 
  labs(
    x = "Value of all economic losses (thousands of US dollars)",
    title = "Distribution of Economic Damage of Disasters from 2000 - 2025",
    subtitle = "With outliers"
  )

summary(events$total_damage_000_us)

events_removedOutliers <- events |>
  filter(as.integer(total_damage_000_us) < (559750 + 1.5* (559750 - 20000)) & 
         as.integer(total_damage_000_us) > (20000 - 1.5 * (559750 - 20000)))

events_removedOutliers |> #without outliers
  ggplot(aes(x = total_damage_000_us)) +
  geom_histogram() + 
  labs(
    x = "Value of all economic losses (thousands of US dollars)",
    title = "Distribution of Economic Damage of Disasters from 2000 - 2025",
    subtitle = "Without outliers"
  )

```

The distribution is right-skewed and unimodal. The center is best represented by the median, which is \$109,576,000, and the spread is best represented with the IQR, which is \$539,750,000 (559,750,000 - 20,000,000). There are 123 (828-705) outliers present in the upper end of our distribution.

Predictors of interests:

1.  disaster_type - the type of disaster (categorical)

    ```         
    ```

2.  country (might change to continent) - country where the disaster occurred and had an impact. If multiple countries are affected, each will have an entry (categorical)

3.  duration - duration in days of the disaster (quantitative)

4.  total_deaths - Total fatalities (deceased and missing combined) (quantitative)

    ```{r}
    #| label: total-deaths-eda

    summary_table <- tibble(
      Statistic = c("Minimum", "1st Quartile", "Median", "Mean", "3rd Quartile", "Maximum"),
      Value = summary(events$total_deaths)
    )

    summary_table |>
      kable(caption = "Summary Statistics (total_deaths)")

    events |>
      ggplot(aes(x = log(total_deaths + 1))) +
      geom_vline(
        xintercept = median(log(events$total_deaths + 1)),
        linetype = "longdash",
        color = "black"
      ) +
      geom_density(fill = "blue", alpha = 0.6) +
      labs(
        title = "Distribution of Log-Transformed Total Deaths",
        x = "Number of Deaths (log transformed)",
        y = "Density"
      ) +
      scale_x_continuous(breaks = seq(0, 20, by = 2))
    ```

    ```{r}
    #| label: total-deaths-eda-bivariate

    events |>
      ggplot(aes(x = log(total_deaths + 1), y = log(total_damage_000_us + 1))) +
      geom_point(alpha = 0.4, color = "blue") +
      geom_smooth(method = "lm", se = TRUE, color = "black") +
      labs(
        title = "Economic Damage vs Total Deaths",
        x = "Total Deaths (log transformed)",
        y = "Economic Damage (thousands USD, log transformed)"
      ) +
      theme_minimal()
    ```

5.  no_injured - Number of people with physical injuries, trauma, or illness requiring immediate medical assistance due to the disaster (quantitative)

    ```{r}
    #| label: number-injured-eda
    #| 
    summary_table <- tibble(
      Statistic = c("Minimum", "1st Quartile", "Median", "Mean", "3rd Quartile", "Maximum"),
      Value = summary(events$no_injured)
    )

    summary_table |>
      kable(caption = "Summary Statistics (no_injured)")

    events |>
      ggplot(aes(x = log(no_injured + 1))) +
      geom_vline(
        xintercept = median(log(events$no_injured + 1)),
        linetype = "longdash",
        color = "black"
      ) +
      geom_density(fill = "lightblue", alpha = 0.6) +
      labs(
        title = "Distribution of Log-Transformed Number of People Injured",
        x = "Number of People Injured",
        y = "Density"
      ) +
      scale_x_continuous(breaks = seq(0, 12, by = 2))

    ```

6.  no_affected - Number of people requiring immediate assistance due to the disaster (quantitative)

    ```{r}
    #| label: number-affected-eda

    summary_table <- tibble(
      Statistic = c("Minimum", "1st Quartile", "Median", "Mean", "3rd Quartile", "Maximum"),
      Value = summary(events$no_affected)
    )

    summary_table |>
      kable(caption = "Summary Statistics (no_affected)")

    events |>
      ggplot(aes(x = log(no_affected + 1))) +
      geom_vline(
        xintercept = median(log(events$no_affected + 1)),
        linetype = "longdash",
        color = "black"
      ) +
      geom_density(fill = "turquoise", alpha = 0.6) +
      labs(
        title = "Distribution of Log-Transformed Number of People Affected",
        x = "Number of People Affected",
        y = "Density"
      ) +
      scale_x_continuous(breaks = seq(0, 100, by = 2))

    ```

    ```{r}
    #| label: number-affected-eda-bivariate

    events |>
      ggplot(aes(x = log(no_affected + 1), y = log(total_damage_000_us + 1))) +
      geom_point(alpha = 0.4, color = "turquoise") +
      geom_smooth(method = "lm", se = TRUE, color = "black") +
      labs(
        title = "Economic Damage vs Number of People Affected",
        x = "Number of People Affeced (log transformed)",
        y = "Economic Damage (thousands USD, log transformed)"
      ) +
      theme_minimal()
    ```

::: callout-important
Before you submit, make sure your code chunks are turned off with `echo: false` and there are no warnings or messages with `warning: false` and `message: false` in the YAML.
:::
