---
title: "Add project title"
author: "Max Niu"
date: "Update date"
format: pdf
execute: 
  warning: false
  message: false
  echo: false
editor: visual
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: load-pkg-data


library(tidyverse)
library(tidymodels)
library(readxl)
library(janitor)
library(lubridate)

raw_events <- read_excel("data/emdat_disasters_2000_2025.xlsx") 

```

## Data processing

```{r}
#| echo: false
#| message: false
#| warning: false

events <- raw_events |>
  clean_names()

events <- events |>
  select(disaster_type, country, start_year, start_month, start_day, end_year, end_month, 
         end_day, total_deaths, no_injured, no_affected, total_affected, total_damage_000_us)


events <- events |>
  drop_na() #selecting was done before because otherwise too many rows would be dropped

events <- events |>
  mutate(
    start_date = make_date(start_year, start_month, start_day),
    end_date = make_date(end_year, end_month, end_day),
    duration_days = as.numeric(end_date - start_date)
  )

events <- events |>
  mutate(
    duration_days = duration_days + 1 # we are including the start date
  )
```

## Introduction

#### Background:

As global temperatures rise, climate-related disasters have increased in both frequency and severity over the past several decades ([World Wild Life](https://www.worldwildlife.org/resources/explainers/is-climate-change-increasing-the-risk-of-disasters/)). Extreme weather events such as floods, droughts, hurricanes, etc. impose a huge human and real economic costs across the world. It is imperative that we understand the determinants of disaster severity (in terms of economic damages) for better policy planning and mitigative strategies.

The Emergency Events Database (EM-DAT), maintained by the Centre for Research on the Epidemiology of Disasters (CRED), keeps an internationally standardized data on natural diasters worldwide. The information includes disaster type, location, affected population, deaths, injuries, and economic damages.

#### Research Question:

How do different disaster characteristics (such as geographic location, time of year, type, and total casualties incurred) predict the total economic damage caused by natural disasters in USD?

-   Specifically, we examine whether disaster type, location, timing, duration, and measures of human impact (deaths, injuries, and affected population) are associated with higher reported economic losses.

#### Motivation:

We were first drawn to this research question because were intrigued by this data; we were amazed by the granularity and the surprising breadth of information in this data. However, we realized that doing research on this matter could have real, practial uses.

First, identifying predictors of large economic damage is critical for both public and private institutions. Governments and international organizations allocate substantial resources towards mitigation and recovery efforts, yet the determinants of high-cost disasters vary across event types and regions. Same applies for insurance companies and hedge fund traders that try modeling key determinants relevant to a disaster for their comparative advavantage against peers in the market.

By analyzing global disaster data, this project aims to quantify which observable characteristics are most strongly associated with economic loss. We hope that these results may provide insight into which types of disasters and impact measures are most economically destructive.

#### Hypothesis Reasoning:

Based on economic reasoning and relevant research, we expect that...

1.  Disasters with higher number of affected individuals (absolute number, number of deaths and injuries) will have a positive relationship with economic damage.
2.  Certain disaster types will incur systematically higher economic damage than that of others.
3.  The level of damage dollars will defer based on countries because of their respective infrastructure that exists to mitigate damage.

#### Data Description:

This data was obtained from the Emergency Events Database, otherwise known as [EM-DAT](https://doc.emdat.be/docs/data-structure-and-content/emdat-public-table/#column-description), which is maintained by the Centre for Research on the Epidemiology of Disasters (CRED). The goal of the database is to compile standardized information on natural disasters that occur around the world. According to the information on metadata, an event is included in EM-DAT if one or more of the following criteria is met:

-   10+ people killed
-   100+ people affected
-   Declaration of a state of emergency
-   Solicitation of international assistance

The observations describe different disasters that occurred around the world. Of each disaster, data on the number of people injured, affected, homeless, or killed, the total economic damages, type of disaster, exact location, start date and end date of the disaster were collected.

The data dictionary can be found [here](https://github.com/sta210-sp26/project-MaxTGH-hyeonsuan06/blob/main/data/README.md).

## Exploratory Data Analysis

```{r}
events |> 
  ggplot(aes(x = total_damage_000_us)) +
  geom_histogram() + 
  labs(
    x = "Value of all economic losses (thousands of US dollars)",
    title = "Distribution of Economic Damage of Disasters from 2000 - 2025",
    subtitle = "With outliers"
  )

summary(events$total_damage_000_us)


events |> #with outliers
  ggplot(aes(x = log(total_damage_000_us))) +
  geom_histogram() + 
  labs(
    x = "Value of all economic losses (thousands of US dollars)",
    title = "Distribution of Economic Damage of Disasters from 2000 - 2025"
  )

events <- events|>
  mutate(log_total_damage_000_us = log(total_damage_000_us))


summary(events$log_total_damage_000_us)
sd(events$log_total_damage_000_us)

```

The distribution of the unchanged response variable is right-skewed and unimodal. The center is best represented by the median, which is \$109,576,000, and the spread is best represented with the IQR, which is \$539,750,000 (559,750,000 - 20,000,000). There are 123 (828-705) outliers present in the upper end of our distribution.

The distribution of the logged response variable is unimodal and symmetric. The center is best represented by the mean, which is \$98,027,173.52 (e\^11.493 \* 1000), and the spread is best represented with the standard deviation, which is \$13544.98 (e\^2.606016 \* 1000). There are outliers present in our distribution.

Predictors of interests:

1.  disaster_type - the type of disaster (categorical)

    ```{r}

    events |>
      distinct(disaster_type)

    events |>
      ggplot(aes(x = disaster_type)) +
      geom_bar() + 
      labs(
        x = "Disaster Type",
        y = "Count",
        title = "Number of Occurences of each Disaster Type"
      )

    ```

    Storms, floods, and earthquakes are the most common disaster types, composing most of the data. Storms are the most common. Extreme temperature, wet mass movements, volcanic activity, and wildfires are the least common disaster types, with extreme temperature being the least common.

    ```{r}

    events |>
      ggplot(aes(x = disaster_type, y = log(total_damage_000_us))) +
      geom_boxplot() + 
      labs(
        x = "Disaster Type",
        y = "Log of Total Economic Damage (Thousands of USD)",
        title = "Distribution of Log of Total Economic Damage for each Disaster Type"
      )
    ```

2.  duration - duration in days of the disaster (quantitative)

    ```{r}

    events |>
      ggplot(aes(x = duration_days)) +
      geom_histogram() +
      labs(
        x = "Duration (Days)",
        y = "Count",
        title = "Distribution of Days of Disasters"
      )

    events |>
      ggplot(aes(x = log(duration_days))) +
      geom_histogram(bins = 15) +
      labs(
        x = "Log of Duration (Days)",
        y = "Count",
        title = "Distribution of Duration of Disasters"
      )
    ```

    ```{r}

    events |>
      ggplot(aes(x = log(duration_days), y = log(total_damage_000_us), color = disaster_type)) +
      geom_point() +
      geom_smooth(method = "lm", se = FALSE) +
      labs(
        x = "Log of Duration in Days",
        y = "Log of Total Economic Damage (Thousands of USD)",
        title = "Relationship between Duration of Natural Disaster and Total Economic Damage",
        subtitle = "Both Variables Log-transformed"
      )

    events |>
      ggplot(aes(x = log(duration_days), y = log(total_damage_000_us))) +
      geom_point() +
      facet_wrap(~disaster_type)+
      geom_smooth(method = "lm", se = FALSE) + 
      labs(
        x = "Log of Duration in Days",
        y = "Log of Total Economic Damage (Thousands of USD)",
        title = "Relationship between Duration of Natural Disaster and Total Economic Damage",
        subtitle = "Both Variables Log-transformed"
      )


    ```

3.  total_deaths - Total fatalities (deceased and missing combined) (quantitative)

4.  no_injured - Number of people with physical injuries, trauma, or illness requiring immediate medical assistance due to the disaster (quantitative)

5.  no_affected - Number of people requiring immediate assistance due to the disaster (quantitative)

::: callout-important
Before you submit, make sure your code chunks are turned off with `echo: false` and there are no warnings or messages with `warning: false` and `message: false` in the YAML.
:::
