---
title: "Add project title"
author: "Add name(s)"
date: "Update date"
format: pdf
execute: 
  warning: false
  message: false
  echo: false
editor: visual
---

```{r}
#| label: load-pkg-data

raw_events <- read_excel("data/emdat_disasters_2000_2025.xlsx") 

```

## Data processing

```{r}
events <- raw_events |>
  clean_names()

events <- events |>
  select(disaster_type, country, start_year, start_month, start_day, end_year, end_month, 
         end_day, total_deaths, no_injured, no_affected, total_affected, total_damage_000_us)


events <- events |>
  drop_na() #selecting was done before because otherwise too many rows would be dropped

events <- events |>
  mutate(
    start_date = make_date(start_year, start_month, start_day),
    end_date = make_date(end_year, end_month, end_day),
    duration_days = as.numeric(end_date - start_date)
  )

```

## Exploratory Data Analysis

```{r}
events |> #with outliers
  ggplot(aes(x = total_damage_000_us)) +
  geom_histogram() + 
  labs(
    x = "Value of all economic losses (thousands of US dollars)",
    title = "Distribution of Economic Damage of Disasters from 2000 - 2025",
    subtitle = "With outliers"
  )

summary(events$total_damage_000_us)

events_removedOutliers <- events |>
  filter(as.integer(total_damage_000_us) < (559750 + 1.5* (559750 - 20000)) & 
         as.integer(total_damage_000_us) > (20000 - 1.5 * (559750 - 20000)))

events_removedOutliers |> #without outliers
  ggplot(aes(x = total_damage_000_us)) +
  geom_histogram() + 
  labs(
    x = "Value of all economic losses (thousands of US dollars)",
    title = "Distribution of Economic Damage of Disasters from 2000 - 2025",
    subtitle = "Without outliers"
  )

```

Predictors of interests (Data from EM-DAT):

1.  disaster_type - the type of disaster (categorical)

2.  country (might change to continent) - country where the disaster occurred and had an impact. If multiple countries are affected, each will have an entry (categorical)

3.  duration - duration in days of the disaster (quantitative)

4.  total_deaths - Total fatalities (deceased and missing combined) (quantitative)

5.  no_injured - Number of people with physical injuries, trauma, or illness requiring immediate medical assistance due to the disaster (quantitative)

6.  no_affected - Number of people requiring immediate assistance due to the disaster (quantitative)

::: callout-important
Before you submit, make sure your code chunks are turned off with `echo: false` and there are no warnings or messages with `warning: false` and `message: false` in the YAML.
:::
